---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Firstly load the data from the zip file:  
- Set the working directory 
- Create a connection to the zip file  
- Load data from activity.csv into activity variable  

```{r results='hide'}
library(dplyr)
library(lubridate)
library(ggplot2)
setwd("~/Desktop/coursera/Assignments/REPRES/RepData_PeerAssessment1")
activity <- read.csv (unz("activity.zip","activity.csv"))
```

Next explore the basic features:
```{r}
str(activity)
```
```{r}
head(activity)
```



Recast the date variable as a date:
```{r}
activity <- mutate(activity, date = ymd(date))
```

## What is mean total number of steps taken per day?
Calculate the mean total number of steps taken per day by firstly grouping by day. Ignore any NA values.
```{r}
total_steps_per_day <- activity %>% group_by(date) %>% summarize(total_steps=sum(steps, na.rm=TRUE))

```

Now plot the histogram of total steps over the measurement period changing the bin width from 30 to 100:

```{r}

qplot(x=total_steps,data=total_steps_per_day, binwidth=1000)

```

Calculate the mean and median of the total_steps:
```{r}
mean_steps <- mean(total_steps_per_day$total_steps)
median_steps <- median(total_steps_per_day$total_steps)

```

The mean is `r prettyNum(mean_steps, big.mark=",")` and the median is `r prettyNum(median_steps, big.mark=",")`.


## What is the average daily activity pattern?
Next group by the interval and make a line plot of the average steps taken cross all days in that interval.

```{r}
average_interval <- activity %>% group_by(interval) %>% summarize(average_steps=mean(steps, na.rm=TRUE))

qplot(x=interval, y=average_steps, data=average_interval, geom="path" )

```


Next the maximum steps:

```{r}
max_steps <-average_interval[which.max(average_interval$average_steps),]

```

The interval with the maximum steps on average is `r max_steps$interval`  


## Imputing missing values
Firstly let's calculate the number of NAs
```{r}
print(sum(is.na(activity$steps)))
```

Fill the NAs with the mean for the equivalent 5 minute interval and add to a new dataset.  
```{r}

completedata <- activity %>% group_by(interval)
completedata <- mutate(completedata, steps= replace(steps,list=is.na(steps), values=mean(steps, na.rm=TRUE)))

```

Make a new histogram of the total steps for the complete data.  
```{r}
completedataplot <- completedata %>% group_by(date) %>% summarize(total_steps=sum(steps))
qplot(x=total_steps,data=completedataplot, binwidth=1000)

```


The new mean and median are:  
```{r}
new_mean_steps <- mean(completedataplot$total_steps)
new_median_steps <- median(completedataplot$total_steps)

```

New mean = `r prettyNum(new_mean_steps, big.mark=",")`  
New median = `r prettyNum(new_median_steps, big.mark=",")`

The impact on each of the removal of missing data is as follows  

Difference to mean = `r prettyNum(new_mean_steps - mean_steps, big.mark=",")`  
Difference to median = `r prettyNum(new_median_steps - median_steps, big.mark=",")`  

## Are there differences in activity patterns between weekdays and weekends?

Add the new factor variable to completedata  
```{r}
completedata$weekday <- as.factor(ifelse(wday(completedata$date)==1|wday(completedata$date)==7,"weekend","week"))

```

and do a panel plot of weekend v weekday  

```{r}
new_average_interval <- completedata %>% group_by(weekday, interval) %>% summarize(average_steps=mean(steps))

qplot(x=interval, y=average_steps, data=new_average_interval, geom="path", facets = weekday~. )
```

